---
title: "Cox Survival Analysis for LGG Samples"
author: "Run Jin"
date: "8/12/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(survival)
library(caret)
library(cowplot)
```

## Import relevant data & select cohort of interest
```{r import data}
histology_df <- read_tsv("../data/histologies.tsv")
expression_data <- readRDS("../data/gene-expression-rsem-tpm-collapsed.rds")

cohort_df <- histology_df %>% dplyr::filter(tumor_descriptor == "Initial CNS Tumor") %>%
  dplyr::filter(experimental_strategy=="RNA-Seq") %>% 
  dplyr::filter(cancer_group %in% c("LGG", "Low-grade glioma/astrocytoma")) %>%
  dplyr::select(Kids_First_Biospecimen_ID, OS_status, OS_days, harmonized_diagnosis, CNS_region)

nrow(cohort_df)
cohort_df
```

## Merge expression data to histology file
```{r sample grouping}
# get biospecimen ID's for samples 
cohort_bsid <- cohort_df %>% pull(Kids_First_Biospecimen_ID) %>% unique()

# subset to gene and sample of interest          
expression_of_interest <- expression_data %>% dplyr::select(all_of(cohort_bsid)) %>% 
  rownames_to_column("GeneSymbol") %>% 
  dplyr::filter(GeneSymbol %in% c("SLC7A5", "FOLH1")) %>% 
  column_to_rownames("GeneSymbol") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("Kids_First_Biospecimen_ID")

#annotate relevant clinical information to the expression data frame
combined <- expression_of_interest %>% dplyr::left_join(cohort_df) 
```

## Modify histology file to mark living as 0 and deceased as 1
```{r os status leveling}
combined_annotated <- combined %>% dplyr::filter(!is.na(OS_status)) %>%
  dplyr::mutate(os_status_level = case_when(
    OS_status == "LIVING" ~ 0,
    OS_status == "DECEASED" ~ 1
  )) 

# change the columns to desired type 
combined_annotated$harmonized_diagnosis <- as.factor(combined_annotated$harmonized_diagnosis)
combined_annotated$CNS_region <- as.factor(combined_annotated$CNS_region)

#the number of samples that HAVE OS_status is:
nrow(combined_annotated)
```
## Build up the model for cox analysis below 
```{r generating models}
model_slc7a5 <- as.formula(survival::Surv(OS_days, os_status_level) ~ harmonized_diagnosis + CNS_region + SLC7A5)
model_folh1 <- as.formula(survival::Surv(OS_days, os_status_level) ~ harmonized_diagnosis + CNS_region + FOLH1)
```

## Create train test partition
```{r train test partition}
set.seed(236)
trainIndex <- createDataPartition(combined_annotated$OS_status, p = 0.6, 
                                      list = FALSE, 
                                      times = 1)
data_train <- combined_annotated[trainIndex, ]
data_test <- combined_annotated[-trainIndex, ]
```

## Run Cox Analysis on training set

First run it for SLC7A5
```{r coxph training 1}
coxph_train_slc7a5 <- survival::coxph(model_slc7a5,
                     data = data_train)

coxph_train_slc7a5_sum <- as.data.frame(summary(coxph_train_slc7a5)$coefficients, keep.rownames = T) %>% 
  tibble::rownames_to_column()
coxph_train_slc7a5_sum 
write_tsv(coxph_train_slc7a5_sum , "results/coxph_train_slc7a5_sum.tsv")
```

Then run it for FOLH1
```{r coxph training 2}
coxph_train_folh1 <- survival::coxph(model_folh1,
                     data = data_train)

coxph_train_folh1_sum <- as.data.frame(summary(coxph_train_folh1)$coefficients, keep.rownames = T) %>% 
  tibble::rownames_to_column()
coxph_train_folh1_sum 
write_tsv(coxph_train_folh1_sum , "results/coxph_train_folh1_sum.tsv")
```

## Calculate risk score for training set and predict for test set 

First calculate for SLC7A5
```{r risk score calculation 1}
train_risk_slc7a5 <- predict(coxph_train_slc7a5, type = "risk")
med_risk_slc7a5 <- median(train_risk_slc7a5)
med_risk_slc7a5

test_risk_slc7a5 <- predict(coxph_train_slc7a5, newdata = data_test, type = "risk") %>% 
  as.data.frame() %>% cbind(data_test) %>%
  dplyr::rename("RiskScore" = ".") %>% 
  dplyr::mutate(RiskGroup = case_when(
    RiskScore > med_risk_slc7a5 ~ "High", 
    RiskScore <= med_risk_slc7a5 ~ "Low"
  ))

test_risk_slc7a5$RiskGroup <- as.factor(test_risk_slc7a5$RiskGroup)
```

Then calculate for FOLH1
```{r risk score calculation 2}
train_risk_folh1 <- predict(coxph_train_folh1, type = "risk")
med_risk_folh1 <- median(train_risk_folh1)
med_risk_folh1

test_risk_folh1 <- predict(coxph_train_folh1, newdata = data_test, type = "risk") %>% 
  as.data.frame() %>% cbind(data_test) %>%
  dplyr::rename("RiskScore" = ".") %>% 
  dplyr::mutate(RiskGroup = case_when(
    RiskScore > med_risk_folh1 ~ "High", 
    RiskScore <= med_risk_folh1 ~ "Low"
  ))

test_risk_folh1$RiskGroup <- as.factor(test_risk_folh1$RiskGroup)
```

## Now run Cox analyses per gene for RiskScore and RiskGroup

First calculate per risk group for SLC7A5
```{r Cox analyses with riskgroup 1}
coxph_riskgroup_slc7a5 <- survival::coxph(formula = Surv(OS_days, os_status_level) ~  RiskGroup,
                     data = test_risk_slc7a5)
coxph_riskgroup_slc7a5_sum <- as.data.frame(summary(coxph_riskgroup_slc7a5)$coefficients) %>% 
  rownames_to_column()
coxph_riskgroup_slc7a5_sum %>% knitr::kable()
```

Calculate per risk score for SLC7A5
```{r Cox analyses with riskscore 1}
coxph_riskscore_slc7a5 <- survival::coxph(formula = Surv(OS_days, os_status_level) ~  RiskScore,
                     data = test_risk_slc7a5)
coxph_riskscore_slc7a5_sum <- as.data.frame(summary(coxph_riskscore_slc7a5)$coefficients) %>% 
  rownames_to_column()
coxph_riskscore_slc7a5_sum %>% knitr::kable()
```

Saving results for SLC7A5 results:
```{r save results 1}
write_tsv(coxph_riskgroup_slc7a5_sum, "results/coxph_test_riskgroup_slc7a5_sum.tsv")
write_tsv(coxph_riskscore_slc7a5_sum, "results/coxph_test_riskscore_slc7a5_sum.tsv")
```

Then calculate per risk group for FOLH1
```{r Cox analyses with riskgroup 2}
coxph_riskgroup_folh1 <- survival::coxph(formula = Surv(OS_days, os_status_level) ~  RiskGroup,
                     data = test_risk_folh1)
coxph_riskgroup_folh1_sum <- as.data.frame(summary(coxph_riskgroup_folh1)$coefficients) %>% 
  rownames_to_column()
coxph_riskgroup_folh1_sum %>% knitr::kable()
```

Calculate per risk score for FOLH1
```{r Cox analyses with riskscore 2}
coxph_riskscore_folh1 <- survival::coxph(formula = Surv(OS_days, os_status_level) ~  RiskScore,
                     data = test_risk_folh1)
coxph_riskscore_folh1_sum <- as.data.frame(summary(coxph_riskscore_folh1)$coefficients) %>% 
  rownames_to_column()
coxph_riskscore_folh1_sum %>% knitr::kable()
```

Saving results for FOLH1 results:
```{r save results 2}
write_tsv(coxph_riskgroup_folh1_sum, "results/coxph_test_riskgroup_folh1_sum.tsv")
write_tsv(coxph_riskscore_folh1_sum, "results/coxph_test_riskscore_folh1_sum.tsv")
```

## Finally plot the survival for each gene per risk group

First generate plot for SLC7A5
```{r plotting slc7c5 risk group survival}
fit_slc7c5 <- survfit(Surv(OS_days, os_status_level) ~ RiskGroup, data = test_risk_slc7a5)
plot_coxph_riskgroup_slc7a5 <- survminer::ggsurvplot(fit_slc7c5,
           data=test_risk_slc7a5,                                
           pval = TRUE, 
           conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           linetype = "strata", # Change line type by groups
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), # Change ggplot2 theme
           palette = c("#E7B800", "#2E9FDF"))

# Make this plot a combined plot
surv_plot_slc7a5 <- cowplot::plot_grid(plot_coxph_riskgroup_slc7a5[[1]], 
                                       plot_coxph_riskgroup_slc7a5[[2]], 
                                       nrow = 2, 
                                       rel_heights = c(2.5, 1))
# Print it out here
surv_plot_slc7a5

# Save the plot
cowplot::save_plot(filename = "plots/coxph_riskgroup_slc7a5_survival.png", plot = surv_plot_slc7a5)
```

Second generate plot for FOLH1
```{r plotting folh1 risk group survival}
fit_folh1 <- survfit(Surv(OS_days, os_status_level) ~ RiskGroup, data = test_risk_folh1)
plot_coxph_riskgroup_folh1 <- survminer::ggsurvplot(fit_folh1,
           data=test_risk_folh1,                                
           pval = TRUE, 
           conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           linetype = "strata", # Change line type by groups
           surv.median.line = "hv", # Specify median survival
           ggtheme = theme_bw(), # Change ggplot2 theme
           palette = c("#E7B800", "#2E9FDF"))

# Make this plot a combined plot
surv_plot_folh1 <- cowplot::plot_grid(plot_coxph_riskgroup_folh1[[1]], 
                                      plot_coxph_riskgroup_folh1[[2]], 
                                      nrow = 2, 
                                      rel_heights = c(2.5, 1))
# Print it out here
surv_plot_folh1

# Save the plot
cowplot::save_plot(filename = "plots/coxph_riskgroup_folh1_survival.png", plot = surv_plot_folh1)
```


