---
title: "Kaplan-Meier Survival Analyses per Gene Expression Level"
author: "Run Jin"
date: "8/11/2021"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(survminer)
library(cowplot)
library(survival)
```

## Import relevant data & select cohort of interest
```{r import data}
histology_df <- read_tsv("../data/histologies.tsv")
expression_data <- readRDS("../data/gene-expression-rsem-tpm-collapsed.rds")

cohort_df <- histology_df %>% dplyr::filter(tumor_descriptor == "Initial CNS Tumor") %>%
  dplyr::filter(experimental_strategy=="RNA-Seq") %>% 
  dplyr::filter(cancer_group %in% c("LGG", "Low-grade glioma/astrocytoma")) %>%
  dplyr::select(Kids_First_Biospecimen_ID, OS_status, OS_days)

nrow(cohort_df)
cohort_df
```

## Separate samples into high vs. low expression group
```{r sample grouping}
# get biospecimen ID's for samples 
cohort_bsid <- cohort_df %>% pull(Kids_First_Biospecimen_ID) %>% unique()

# subset to gene and sample of interest          
expression_of_interest <- expression_data %>% dplyr::select(all_of(cohort_bsid)) %>% 
  rownames_to_column("GeneSymbol") %>% 
  dplyr::filter(GeneSymbol %in% c("SLC7A5", "FOLH1")) %>% 
  column_to_rownames("GeneSymbol") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("Kids_First_Biospecimen_ID")

#annotate relevant clinical information to the expression data frame
combined <- expression_of_interest %>% dplyr::left_join(cohort_df) 

#for genes of interest, generate grouping based on that gene
#define the number we want to slice
n=0.5*nrow(combined)

high_exp_SLC7A5 <- combined %>% 
  top_n(n, wt=SLC7A5) %>% 
  mutate(expression_group_SLC7A5 = "High")
low_exp_SLC7A5 <- combined %>% 
  top_n(-n, wt=SLC7A5) %>% 
  mutate(expression_group_SLC7A5 = "Low")

combined_SLC7A5 <- rbind(high_exp_SLC7A5, low_exp_SLC7A5)

high_exp_FOLH1 <- combined %>% 
  top_n(n-1, wt=FOLH1) %>% #this is needed since samples in the border would be counted twice otherwise
  mutate(expression_group_FOLH1 = "High")
low_exp_FOLH1 <- combined %>% 
  top_n(-n, wt=FOLH1) %>% 
  mutate(expression_group_FOLH1 = "Low")
combined_FOLH1 <- rbind(high_exp_FOLH1, low_exp_FOLH1)

# combine annotations with expression data together
combined_annotated <- left_join(combined_SLC7A5, combined_FOLH1)
```

## Modify histology file to mark living as 0 and deceased as 1
```{r os status leveling}
combined_annotated <- combined_annotated %>% dplyr::filter(!is.na(OS_status)) %>%
  dplyr::mutate(os_status_level = case_when(
    OS_status == "LIVING" ~ 1,
    OS_status == "DECEASED" ~ 0
  )) 

#the number of samples that HAVE OS_status is:
nrow(combined_annotated)
```

## Build up the model formula for both statistical testing below 
```{r generating models}
model_slc7a5 <- "survival::Surv(OS_days, os_status_level) ~ expression_group_SLC7A5"
model_folh1 <- "survival::Surv(OS_days, os_status_level) ~ expression_group_FOLH1"

```

## Running Log-Rank Analyses 

For testing whether two survival curves are different, we can use Log-Rank which
can be implemented using `survival::survdiff` function. 
In this example, we will test for survival curve differences between for 
`expression_level_SLC7A5` or `expression_level_FOLH1` categories. 
Use the `survdiff` function to test the differences between the `High` and `Low` 
curves.

```{r Do log rank test}
fit_slc7a5 <- survival::survdiff(
      as.formula(model_slc7a5),
      data = combined_annotated
    )

fit_folh1 <- survival::survdiff(
      as.formula(model_folh1),
      data = combined_annotated
    )

# Obtain p value for Chi-Squared stat
fit_slc7a5$p.value <- pchisq(fit_slc7a5$chisq, df = 1, lower = FALSE)
fit_folh1$p.value <- pchisq(fit_folh1$chisq, df = 1, lower = FALSE)
```

## Printing table of Log-Rank Results

First print out SLC7A5 results
```{r Make log rank table 1}
# Print out the table here
fit_slc7a5_df <- as.data.frame(fit_slc7a5[c("n", "obs", "exp", "chisq", "p.value")])
fit_slc7a5_df %>% knitr::kable()
```

Second print out FOLH1 results
```{r Make log rank table 2}
# Print out the table here
fit_folh1_df <- as.data.frame(fit_folh1[c("n", "obs", "exp", "chisq", "p.value")])
fit_folh1_df %>% knitr::kable()
```

Now save the results as tables
```{r Write log rank to table}
# Save the table data in a TSV
readr::write_tsv(fit_slc7a5_df, "results/log_rank_survival_slc7a5.tsv")
readr::write_tsv(fit_folh1_df, "results/log_rank_survival_folh1.tsv")
```

## Running Kaplan-Meier Fit

The [Kaplan-Meier](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3059453/) 
survival curve shows the probability of survival over time.

```{r Kaplan Meier}
kap_fit_slc7a5 <- survival::survfit(
      as.formula(model_slc7a5),
      data = combined_annotated
    )

kap_fit_folh1 <- survival::survfit(
      as.formula(model_folh1),
      data = combined_annotated
    )
```

`survminer` package comes with a convenient plotting function which uses `ggplot2`
arguments. 
It requires the original Kaplain-Meier `survfit` object. In here, 

First let's plot for SLC7A5:
```{r Make survival plot 1}
surv_plot <- survminer::ggsurvplot(kap_fit_slc7a5,
  pval = fit_slc7a5$p.value, # use the computed pval from log rank analyses 
  data = combined_annotated,
  risk.table = TRUE,
  xlim = c(0, 2000),
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE
)
# Make this plot a combined plot
surv_plot <- cowplot::plot_grid(surv_plot[[1]], surv_plot[[2]], nrow = 2, 
                                rel_heights = c(2.5, 1))
# Print it out here
surv_plot
```

Save the plot to a file. 

```{r Save survival plot 1}
# We can save the plot like a normal ggplot
cowplot::save_plot(filename = "plots/kaplan_meier_survival_slc7a5.png", plot = surv_plot)
```

Second let's plot for FOLH1:
```{r Make survival plot 2}
surv_plot <- survminer::ggsurvplot(kap_fit_folh1,
  pval = fit_folh1$p.value, # use the computed pval from log rank analyses 
  data = combined_annotated,
  risk.table = TRUE,
  xlim = c(0, 2000),
  break.time.by = 500,
  ggtheme = theme_minimal(),
  risk.table.y.text.col = TRUE,
  risk.table.y.text = FALSE
)
# Make this plot a combined plot
surv_plot <- cowplot::plot_grid(surv_plot[[1]], surv_plot[[2]], nrow = 2, 
                                rel_heights = c(2.5, 1))
# Print it out here
surv_plot
```

Save the plot to a file. 

```{r Save survival plot 2}
# We can save the plot like a normal ggplot
cowplot::save_plot(filename = "plots/kaplan_meier_survival_folh1.png", plot = surv_plot)
```

